Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-07-15 manuscript/manuscript.Rmd
incorporates some of Peter's suggestions

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
csl: components/ecology.csl

@@

Introduction
------------
Population models are important tools for predicting the impacts of environmental change on species persistence and abundance. But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012]. The major hurdle is that most population models are built using data from a single study site because collecting those data, which involves tracking the fates of individuals plants, is so difficult. The resulting models cannot be applied to the landscape and regional scales relevant to decision-makers without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007]. The temporal coverage of demographic datasets is no better. [I think the temporal extent is a separate issue, might muddy the waters to bring it up here...] For example, the most common study duration in the COMPADRE matrix population model database is 4 years and only a few studies exceed 10 years [@Salguero-Gomez2015]. The limited spatio-temporal extent of individual-level demographic datasets constrains our ability to use population models to address applied questions about the consequences of climate change.

@@

Aggregate measures of population status, rather than individual performance, offer an intriguing alternative for modeling populations [@Clark2004; @Freckleton2011]. Population-level data cannot provide inference about 
#will never match the precision of individual-level data, IS THIS TRUE? WHAT DOES PRECISION MEAN?
demographic mechanisms, but it might be sufficient for modeling future population states, especially since 
it is feasible to collect across broad spatial extents. The choice between individual and population-level data involves a difficult trade-off: while individual-level data leads to more mechanistic models, population-level data leads to models that can be applied over greater spatial and temporal extents. An open question is how much forecasting skill is lost when we build models based on population rather than individual-level data.

To date, most empirical population modelers have relied on individual-level data, with few attempts to exploit population-level data. An important exception was an effort by @Taylor2004 to model the population growth rate of an invasive species to identify the best strategies for invasion control. They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure. Building on this work, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 demonstrated the application of such methods in a study on arable weeds. In particular, @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models. Thus, population models based on coarse, population-level data show promise for producing ecological forecasts at landscape and regional scales [@Queenborough2011]. However, none of these models included environmental covariates.

Addressing climate change questions with models fit to population-level data is potentially problematic. It is individuals, not populations, that respond to climate variables [@Clark2012]. Ignoring this fact puts us in uneasy proximity to an "ecological fallacy," where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988]. For example, individual plants may respond positively to precipitation but a negative trend is observed at the population level due to increased competition among plants as they grow larger and consume more resources. 
# I think you need a better example. This example makes the population level approach look more sensible than the
# individual level approach. Better to introduce the ideas we focus on in Discussion--growth vs survival vs 
# fecundity, with different signals at different time scales...? 
If population-level data cannot detect important impacts of climate drivers on populations, then population models built with such data will make poor forecasts.

Here, we ask whether statistical and population models based on aggregated, population-level data can detect climate signals as wells as models based on individual-level data. We used a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover. In both models, interannual variation is explained, in part, by climate covariates. We then performed simulations to quantify the sensitivities of species' cover to small perturbations in average precipitation and temperature. We found that population models based on detailed demographic data are more accurate and precise than models based on aggregated data. While both types of models detected climate signals, they produced inconsistent forecasts, in some cases even predicting opposing qualitative responses to a climate perturbation. For these species at this location, detailed demographic data appears necessary to make accurate forecasts. A worrying caveat to our work is that forecasts from both models were very uncertain when we considered full process and parameter uncertainty. Even 14 years worth of demographic data may not be sufficient to make useful forecasts.

@@

From 1932 to 1945 individual plants were identified and mapped annually in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment). In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions. The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [cite Lauenroth and Adler 2008, Chu et al. 2014]. Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.

In this paper, we model populations based on two levels of data: individual and quadrat (Fig. 2). The individual data is the "raw" data. For the quadrat level we data we simply sum individual basal cover for each quadrat by species. This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error. Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.

# might be better to say that we have quantified measurement error and it doesn't seem large enough to
# influence our results (Chu and Adler 2014)

All R code and data necessary to reproduce our analysis is archived on GitHub as release v1.0[^revnote] (http://github.com/atredennick/MicroMesoForecast/releases). That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work.

@@

# You prefer GitHub to a Data Dryad package? Dryad would be more dependable long term...?

@@

# let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...

We fit all models using a hierarchical Bayesian approach. The models are fully descibed in Appendix A, so here we focus on the main process and the model likelihood. For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P for survival, growth, recruitment, or population growth). For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet survives from *t* to *t+1*, or not.

@@

which means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species. Our stastical model for recruitment (_R_, described above) returns the number of new recruit produced per quadrat. Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$) to incorporate the recruitment function in the spatially-implicit IPM. [You probably should say something about spatial structure and cite Adler et al. 2010 EcoLetts for the "no overlap rule"]

@@

#### Quad-based model
Our quad-based model (QBM) perfectly mirrors [I think you need to say a little more about how you simulate it] its statistical description (Eqs. 8-9). We use the same approach for drawing parameter values as described for the IPM.

@@

@@

The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates. Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations. Species showed similar trends (Fig. 3e-h). Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig. 3e-h). *B. gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig. 3e). In isolation, recruitment and survival were insensitive to climate factors for *H. comata* (Fig. 3f). Survival and recruitment of *P. smithii* were both sensitive, negatively, to temperature and precipitation (Fig. 3g). *P. secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig. 3h). The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig. 3h). At least two of three vital rates were sensitive to climate for each species (Fig. 3).

Forecasts based on 1% climate changes were extremely uncertain when we considered model error and parameter uncertainty (Fig. 4). As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P. smithiii* (Fig. 4).

Discussion
----------

MOVE FIRST TWO PARAGRAPHS TO INTRO. MOVE FIG 5 and 6 TO METHODS AND RESULTS

As a potential remedy to the "data dearth" problem, @Queenborough2011 and @Freckleton2011, building on work by @Taylor2004, advocate a "density-structured" modeling approach. Such models do not require individual level demographic data and can adequately describe population dynamics [@Queenborough2011]. The results from density-structured models are not as precise as those from traditional population models, but the loss in precision is traded off with a gain in data. The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks. This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least). The appeal of density-structured approaches is clear.

However, at their core, density-structured models rely on individual level data aggregated to a population level metric (e.g., density classes or percent cover). This creates a potential problem if such models are to be used in a climate change context because inidividuals respond to climate, not populations [@Clark2012]. Are models based on population level metrics as sensitive to climate as models based on individual level metrics? Do these two types of models produce consistent forecasts? Do we need detailed demographic data to forecast the impacts of climate change? These are the questions we sought to answer here.

### The IPM and QBM produced inconsistent forecasts
Using individual and aggregated forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on aggregated data. Our quad-based model (QBM) is based on percent cover data and so is in the spirit of density-structured models. In terms of each model's forecasting ability, the IPM outperformed the QBM (Table 1). This is unsurprising since we expected to lose some precision at the aggregated level. However, the underwhelming performance of the QBM could call into the question forecasts that differ from the IPM.

>>>>

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{todonotes}
bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
csl: components/ecology.csl

@@

Introduction
------------

Population models are important tools for predicting the impacts of environmental change on species persistence and abundance. But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012]. The major hurdle is that most population models are built using data from a single study site because collecting those data, which involves tracking the fates of individuals plants, is so difficult. The resulting models cannot be applied to the landscape and regional scales relevant to decision-makers without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007]. The temporal coverage of demographic datasets is no better. [I think the temporal extent is a separate issue, might muddy the waters to bring it up here...] For example, the most common study duration in the COMPADRE matrix population model database is 4 years and only a few studies exceed 10 years [@Salguero-Gomez2015]. The limited spatio-temporal extent of individual-level demographic datasets constrains our ability to use population models to address applied questions about the consequences of climate change.

@@

Aggregate measures of population status, rather than individual performance, offer an intriguing alternative for modeling populations [@Clark2004; @Freckleton2011]. Population-level data cannot provide inference about demographic mechanisms, but it might be sufficient for modeling future population states, especially since it is feasible to collect across broad spatial extents [e.g., @Queenborough2011]. The choice between individual and population-level data involves a difficult trade-off: while individual-level data leads to more mechanistic models, population-level data leads to models that can be applied over greater spatial and temporal extents. An open question is how much forecasting skill is lost when we build models based on population rather than individual-level data.

To date, most empirical population modelers have relied on individual-level data, with few attempts to exploit population-level data. An important exception was an effort by @Taylor2004 to model the population growth rate of an invasive species to identify the best strategies for invasion control. They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure. Such models do not require individual level demographic data and can adequately describe population dynamics.

Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 demonstrated the application of such methods in a study on arable weeds. In particular, @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models. The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks. This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least). The appeal of density-structured approaches is clear. However, none of these models included environmental covariates.

Addressing climate change questions with models fit to population-level data is potentially problematic. It is individuals, not populations, that respond to climate variables [@Clark2012]. Ignoring this fact puts us in uneasy proximity to an "ecological fallacy", where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988]. Population growth (or decline) is an epiphenomenon of demographic processes like survival, growth, and recruitment that occur at the level of individual plants. Climate can affect each demographic process in unique, potentially opossing, ways [@Dalgleish2011]. These unique climate responses may difficult resolve in statistical models based on population-level data where demographic processes are not identifiable. If population-level data cannot detect important impacts of climate drivers on populations, then population models built with such data will make poor forecasts.

Here, we ask whether statistical and population models based on aggregated, population-level data can detect climate signals as wells as models based on individual-level data. We used a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover. In both models, interannual variation is explained, in part, by climate covariates. We then performed simulations to quantify the sensitivities of species' cover to small perturbations in average precipitation and temperature. We found that population models based on detailed demographic data are more accurate and precise than models based on aggregated data. Our results suggest the popuation-level model is less accurate and precise because important demographic climate signals go undetected. For these species at this location, detailed demographic data appears necessary to make accurate forecasts. A worrying caveat to our work is that forecasts from both models were very uncertain when we considered full process and parameter uncertainty. Even 14 years worth of demographic data may not be sufficient to make useful forecasts.

@@

From 1932 to 1945 individual plants were identified and mapped annually in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment). In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions. The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [@Lauenroth2008; @Chu2014a]. Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.

In this paper, we model populations based on two levels of data: individual and quadrat (Fig. 2). The individual data is the "raw" data. For the quadrat level we data we simply sum individual basal cover for each quadrat by species. This is equivalent to a near-perfect census of quadrat percent cover because previous analysis shows that measurement error at the individual level is small [@Chu2014]. Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.

All R code and data necessary to reproduce our analysis is archived on GitHub as release v1.0[^revnote] (http://github.com/atredennick/MicroMesoForecast/releases). That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work. We have also deposited the v1.0 release on Dryad (*link here after acceptance*).

@@

@@

\todo{let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...}

We fit all models using a hierarchical Bayesian approach. The models are fully descibed in Appendix A, so here we focus on the main process and the model likelihood. For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P$ for survival, growth, recruitment, or population growth). For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet survives from *t* to *t+1*, or not.

@@

which means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species. Our stastical model for recruitment (_R_, described above) returns the number of new recruit produced per quadrat. Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$) to incorporate the recruitment function in the spatially-implicit IPM. \todo{Add spatial structure mention.}

@@

#### Quad-based model
Our quad-based model (QBM) perfectly mirrors its statistical description (Eqs. 8-9). We use the same approach for drawing parameter values as described for the IPM. After drawing the appropriate parameter set, we calculate the mean response (population cover at *t+1* = $x_{t+1}$) according to Eq. 8. We then make a random draw from a [0,1] truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq. 8 and the variance estimate from the fitted model. We can then iterate the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.

@@

Our *a priori* expectation is that the IPM will produce more accurate and precise forecasts. This could be due to differences in sample sizes leading to larger parameter uncertainty for the QBM, or due to the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level. To assess the impact of sample size on QBM parameter uncertainty we refit the QBM statistical model (Eqs. 8-9) after removing sets of 2, 5, 10, and 15 quadrats. We fit 10 models at each level of quadrat removal (2, 5, 10, 15 quadrats), removing a different randomly selected set of quadrats for each fit. We calculated the standard deviation of climate main effects (pptlag, ppt1, ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over replicates within each set of quadrat removals. This allowed us to regress sample size against parameter uncertainty.

To see if the QBM climate effects are correlated, or not, with climate effects for each vital rate model for the IPM, we simply regressed the QBM climate coefficients against each vital rate model's climate coefficients and calculate Pearson's $\rho$. Strong correlations indicate the QBM is capable of detecting climate effects associated with individual vital rates. A weak correlation indicates the QBM "misses" the climate effect on a particular vital rate.

@@

The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates. Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations. Across all species, climate covariates can have opposing effects on different vital rates (Fig. 3e-h).

Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig. 3e-h). *B. gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig. 3e). In isolation, recruitment and survival were insensitive to climate factors for *H. comata* (Fig. 3f). Survival and recruitment of *P. smithii* were both sensitive, negatively, to temperature and precipitation (Fig. 3g). *P. secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig. 3h). The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig. 3h). At least two of three vital rates were sensitive to climate for each species (Fig. 3).

Forecasts based on 1% climate changes were extremely uncertain when we considered model error and parameter uncertainty (Fig. 4). As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P. smithiii* (Fig. 4). IPM forecasts for *P. smithiii* were very uncertain due to overcompensating density dependence and nonlinearities recruitment that get amplified by parameter uncertainty (Appendx C). Neither model was capable of making forecasts of proportional cover change distinguishable from 0 (i.e., no change in population state) when we included model error and parameter uncertainty.

### Sources of uncertainty in the QBM
Sample size had a relatively weak effect on QBM climate parameter uncertainty after the number of quadrats exceeded about 10 (Fig. 5). Inverse-gaussian fits show that increasing sample size beyond the number of quadrats we used will results in deminishing returns in terms of parameter certainty (Fig. 5).

Climate effects estimated from the QBM are most correlated with climate effects from the growth regression at the individual level (Fig. 6). In no case does the QBM statistical model have strong correlations across all three vital rates (Fig. 6). QBM climate effects were most weakly correlated with those from individual-level recruitment models for *B. gracilis*, *H. comata*, and *P. secunda* (Fig. 6a,b,d). For *P. smithii*, QBM climate effects showed no correlation with the survival model effects (Fig. 6c).

Discussion
----------
Population models built using individual-level data allow inference on demographic processes, but their ability to forecast future population states is limited by the spatial extent of the data. Population-level data, like percent cover or discrete density states, are much easier to collect of broad spatial extents, so populations built using such data offer an appealing alternative to traditional population models [@Queenborough2011]. However, at their core, density-structured models rely on individual level data aggregated to a population level metric. This creates a potential problem if such models are to be used in a climate change context because inidividuals respond to climate, not populations [@Clark2012]. Are models based on population level metrics as sensitive to climate as models based on individual level metrics? Do these two types of models produce consistent forecasts? Do we need detailed demographic data to forecast the impacts of climate change? These are the questions we sought to answer here.

### The IPM and QBM produced inconsistent forecasts
Using individual-level and population-level forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on percent cover data. Our quad-based model (QBM) is based on percent cover data and so is in the spirit of density-structured models. In terms of each model's forecasting ability, the IPM outperformed the QBM (Table 1). This is unsurprising since we expected to lose some precision at the aggregated level. However, the underwhelming performance of the QBM could call into the question forecasts that differ from the IPM.
